---
title: "Obj_1_tree_annotations"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Overview

This notebook reads a TreeAnnotator output tree (with all its embedded BEAST
annotations), calibrates the x-axis to calendar time using the most recent
tip's sampling date, joins an external metadata TSV onto the tips, and
produces a time-scaled tree figure with year gridlines and node-age
uncertainty bars.

Edit the **User settings** chunk below, then run the whole document
(`Render` or `Ctrl/Cmd + Shift + K`).

## Setup


```{r}
#| label: load-packages

library(treeio)
library(ggtree)
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(readr)
library(ggplot2)
```

## User settings

Edit these to match your files and analysis.

```{r}
#| label: user-settings

setwd("~/Objective1_BEAST/Objective_1_tree_annotations")
 
tree_file     <- "Data/Clade1_GTR_strict_BICEPS_comb_500_annotated.trees"    # your TreeAnnotator output file

# Your tip labels are structured as: isolate_ID|SBI|Island|Source|date
# List the field names in that exact order here:
tip_label_fields <- c("isolate_ID", "SBI", "Island", "Source", "date")
tip_label_sep    <- "|"

year_gridline_every <- 5           # gridline spacing, in years
output_plot   <- "time_tree.pdf"   # output figure filename
```

## Read the tree

Reads the tree while keeping all BEAST/TreeAnnotator annotations (posterior,
height_95%_HPD, rate, etc.) attached to each node.

```{r}
#| label: read-tree

beast_tree <- read.beast(tree_file)
beast_tree
```

Check which annotation fields are available — useful for confirming the exact
name of the HPD field before plotting (it can appear as `height_0.95_HPD` or
`height_95%_HPD` depending on TreeAnnotator settings/version):

```{r}
#| label: inspect-annotations

names(as_tibble(beast_tree))
```

## Parse metadata out of the tip labels

Since your metadata is already embedded in each tip label
(`isolate_ID|SBI|Island|Source|date`), we split on `|` to recover it as a
proper table, rather than needing a separate TSV. `Name` is kept as the
**full original tip label**, unchanged, so it still matches the tree exactly
for the join later.

```{r}
tree_tips <- beast_tree@phylo$tip.label

metadata <- tibble(Name = tree_tips) |>
  separate_wider_delim(Name, delim = tip_label_sep, names = tip_label_fields,
                        cols_remove = FALSE, too_many = "merge",
                        too_few = "align_start") |>
  mutate(date_parsed = dmy(date)) |>
  relocate(Name)

metadata
```

```{r}
#| label: check-date-parsing

# Flag any tip whose date field failed to parse -- fix these in the
# tip labels (or check for a different date format) before continuing.
metadata |> filter(is.na(date_parsed)) |> select(Name, date)
```

```{r}
#| label: most-recent-tip-date

most_recent_tip_date <- max(metadata$date_parsed, na.rm = TRUE)
most_recent_tip_date
```

This is auto-detected from the `date` field across all tips — double-check
it looks right (i.e. it's genuinely your youngest sample) before moving on.

## Build the plot

`mrsd` ("most recent sampling date") is what calibrates the x-axis to real
calendar time, using the node heights TreeAnnotator already calculated.

```{r}
#| label: build-plot
#| fig-width: 10
#| fig-height: 8

p <- ggtree(beast_tree, mrsd = as.character(most_recent_tip_date),
            ladderize = TRUE) %<+% metadata +   # joins parsed metadata by "Name"/tip label
  theme_tree2() +
  #geom_range(range = "height_0.95_HPD", color = "steelblue", alpha = 0.4, size = 2) +
  geom_tippoint(aes(color = Source), size = 2) +
  geom_tiplab(aes(label = isolate_ID), size = 2.5, offset = 0.3) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
  theme(
    panel.grid.major.x = element_line(color = "grey85", linewidth = 0.3),
    panel.grid.minor.x = element_blank(),
    legend.position = "right"
  ) +
  labs(x = "Year", color = "Source")

p
```

## Save the figure

```{r}
#| label: save-plot

ggsave(output_plot, plot = p, width = 10, height = 8, units = "in")
message("Saved plot to: ", output_plot)
```



