---
title: "Obj_1_tree_annotations"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Overview

This notebook reads a TreeAnnotator output tree (with all its embedded BEAST
annotations), calibrates the x-axis to calendar time using the most recent
tip's sampling date, joins an external metadata TSV onto the tips, and
produces a time-scaled tree figure with year gridlines and node-age
uncertainty bars.

Edit the **User settings** chunk below, then run the whole document
(`Render` or `Ctrl/Cmd + Shift + K`).

## Setup


```{r}
#| label: load-packages

library(treeio)
library(ggtree)
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(readr)
library(ggplot2)
library(scales)
library(ggnewscale) 
library(tibble)
```

## Clade 1


```{r}
#| label: user-settings

setwd("~/Objective1_BEAST/Objective_1_tree_annotations")
 
tree_file     <- "Data/Clade1_GTR_strict_BICEPS_comb_500_annotated.trees"    # your TreeAnnotator output file

# Your tip labels are structured as: isolate_ID|SBI|Island|Source|date
# List the field names in that exact order here:
tip_label_fields <- c("isolate_ID", "SBI", "Island", "Source", "date")
tip_label_sep    <- "|"

year_gridline_every <- 5           # gridline spacing, in years
output_plot   <- "time_tree.pdf"   # output figure filename
```

## Read the tree

Reads the tree while keeping all BEAST/TreeAnnotator annotations (posterior,
height_95%_HPD, rate, etc.) attached to each node.

```{r}
#| label: read-tree

beast_tree <- read.beast(tree_file)
beast_tree
```

Check which annotation fields are available — useful for confirming the exact
name of the HPD field before plotting (it can appear as `height_0.95_HPD` or
`height_95%_HPD` depending on TreeAnnotator settings/version):

```{r}
#| label: inspect-annotations

names(as_tibble(beast_tree))
```

## Parse metadata out of the tip labels

Since your metadata is already embedded in each tip label
(`isolate_ID|SBI|Island|Source|date`), we split on `|` to recover it as a
proper table, rather than needing a separate TSV. `Name` is kept as the
**full original tip label**, unchanged, so it still matches the tree exactly
for the join later.

```{r}
#| label: parse-tip-labels

tree_tips <- beast_tree@phylo$tip.label

metadata <- tibble(Name = tree_tips) |>
  separate_wider_delim(Name, delim = tip_label_sep, names = tip_label_fields,
                        cols_remove = FALSE, too_many = "merge",
                        too_few = "align_start") |>
  mutate(
    # some pipe-delimited fields (e.g. Island) are blank for certain tips --
    # convert empty strings to real NA so they're handled as missing data
    # downstream, rather than an unmatched category that renders grey
    across(all_of(tip_label_fields), ~ na_if(.x, "")),
    # dates in tip labels are dd/mm/yyyy, e.g. "27/04/2014"
    date_parsed = dmy(date)
  ) |>
  relocate(Name)   # %<+% joins on the FIRST column, so Name must lead

metadata


# quick check: how many tips are missing data in each field
metadata |>
  summarise(across(all_of(tip_label_fields), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "field", values_to = "n_missing")
```

```{r}
#| label: check-date-parsing

# Flag any tip whose date field failed to parse -- fix these in the
# tip labels (or check for a different date format) before continuing.
metadata |> filter(is.na(date_parsed)) |> select(Name, date)
```

```{r}
#| label: most-recent-tip-date

most_recent_tip_date <- max(metadata$date_parsed, na.rm = TRUE)
most_recent_tip_date
```

This is auto-detected from the `date` field across all tips — double-check
it looks right (i.e. it's genuinely your youngest sample) before moving on.

## Build the plot

`mrsd` ("most recent sampling date") is what calibrates the x-axis to real
calendar time, using the node heights TreeAnnotator already calculated.



``` {r}
#| label: build-annotated-tree
#| fig-width: 16
#| fig-height: 8

# ---- Base tree, built once ----
p <- ggtree(beast_tree, mrsd = as.character(most_recent_tip_date), ladderize = TRUE) %<+% metadata +
  theme_tree2() +
  # node age uncertainty (95% HPD bars) -- rename if your field is named differently
  geom_range(range = "height_0.95_HPD", color = "steelblue", alpha = 0.4, linewidth = 2) +
  geom_tippoint(aes(color = Source), size = 2) +
  # aligned tip labels with dotted guide lines back to the actual tip position
  geom_tiplab(aes(label = isolate_ID), size = 2, offset = 0.3, align = TRUE, linetype = "dotted") +
  theme(
    panel.grid.major.x = element_line(color = "grey85", linewidth = 0.3),
    panel.grid.minor.x = element_blank(),
    legend.position = "right"
  ) +
  labs(x = "Year", color = "Source")

# ---- Fixed tick/gridline positions, based on the tree's real range only ----
# (a live scales::pretty_breaks() function gets re-evaluated against the
# strip-widened panel later, which is why ticks kept extending outward)
tree_x_breaks <- scales::pretty_breaks(n = 15)(range(p$data$x, na.rm = TRUE))
p <- p + scale_x_continuous(breaks = tree_x_breaks)

# ---- Strip data: one column each, tip labels as row names ----
island_data <- metadata |> select(Name, Island) |> column_to_rownames("Name")
sbi_data    <- metadata |> select(Name, SBI)    |> column_to_rownames("Name")

# ---- Strip data: one column each, tip labels as row names ----
# Island NAs converted to "Unknown" so they can be coloured as an ordinary
# category (see island_colors below). Using the literal string "NA" here
# instead can get silently re-interpreted as a true missing value again by
# later steps, so "Unknown" is the safer, unambiguous choice.
island_data <- metadata |>
  mutate(Island = ifelse(is.na(Island), "Unknown", Island)) |>
  select(Name, Island) |>
  column_to_rownames("Name")
sbi_data    <- metadata |> select(Name, SBI)    |> column_to_rownames("Name")

# ---- Strip sizing, based on the tree's actual year-span ----
tree_width_years  <- diff(range(p$data$x, na.rm = TRUE))
offset1           <- 0.18 * tree_width_years    # space for tip labels before Island strip
strip_width_frac  <- 0.04                       # each strip = 4% of tree width
strip_width_years <- strip_width_frac * tree_width_years
gap_years         <- 0.5                        # gap between the two strips
offset2           <- offset1 + strip_width_years + gap_years

# ---- Custom colour palettes -- edit to match your actual categories ----
island_colors <- c(
  "NI" = "#a6cee3",
  "SI" = "#b2df8a",
  "Unknown"="#ffffff"
)

sbi_colors <- c(
"AY2a"=	"#1F77B4",
"ASY2c"=	"#AEC7E8",
"WY12a"=	"#FF7F0E",
"Y12a"=	"#FFBB78",
"AS2c"=	"#2CA02C",
"SY2c"=	"#98DF8A",
"A2a"=	"#D62728",
"AWY2a"=	"#FF9896",
"Y1a"=	"#9467BD",
"AS12c"=	"#C5B0D5",
"ASWY2c"=	"#8C564B",
"ASY"	="#C49C94",
"ASY12c"=	"#E377C2",
"AWY12a"=	"#F7B6D2",
"AY"=	"#7F7F7F",
"SWY2c"=	"#C7C7C7",
"WY1a"=	"#BCBD22",
"Y2c"	="#DBDB8D"
)

# ---- Strip 1: Island ----
p2 <- p + new_scale_fill()
p2 <- gheatmap(
  p2, island_data,
  offset = offset1, width = strip_width_frac,
  colnames_position = "top", colnames_angle = 90,
  colnames_offset_y = 1.5, font.size = 3
) +
  scale_fill_manual(values = island_colors, name = "Island")

# ---- Strip 2: SBI ----
p2 <- p2 + new_scale_fill()
p2 <- gheatmap(
  p2, sbi_data,
  offset = offset2, width = strip_width_frac,
  colnames_position = "top", colnames_angle = 90,
  colnames_offset_y = 1.5, font.size = 3
) +
  scale_fill_manual(values = sbi_colors, name = "SBI")

# ---- Clamp the displayed x-axis range so it doesn't blow out into the far future ----
p2 <- p2 + coord_cartesian(
  xlim = c(min(p$data$x, na.rm = TRUE) - 5, max(p$data$x, na.rm = TRUE) + 25),
  expand = FALSE
)

p2

```


## Save the figure

```{r}
#| label: save-plot

ggsave(output_plot, plot = p, width = 10, height = 8, units = "in")
message("Saved plot to: ", output_plot)
```


## User settings

```{r}
#| label: skyline-user-settings

skyline_file <- "Data/Clade1_BS_analysis_2014.696.txt"   # your exported Tracer skyline text file
```

## Read the data

The file has a title row before the real header (`time  date  datetime
milliseconds  mean  median  upper  lower`), so we skip the first line.

```{r}
#| label: read-skyline

skyline <- read_tsv(skyline_file, skip = 1, show_col_types = FALSE)
skyline
```

```{r}
#| label: inspect-skyline

glimpse(skyline)
```

## Plot

`time` is already a decimal calendar year (e.g. `2014.696`), so it can be
used directly on the x-axis without further conversion. Population size is
plotted on a log scale, which is standard for skyline plots since Ne can
span orders of magnitude.

```{r}
#| label: plot-skyline
#| fig-width: 9
#| fig-height: 6

p_skyline <- ggplot(skyline, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = median), color = "steelblue4", linewidth = 0.8) +
  scale_y_log10(labels = scales::comma) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 12)) +
  labs(
    x = "Year",
    y = "Effective population size (log scale)",
    title = "Bayesian Skyline Plot"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey88")
  )

p_skyline

p_skyline <- ggplot(skyline, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = median), color = "steelblue4", linewidth = 0.8) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 12)) +
  labs(
    x = "Year",
    y = "Effective population size (Ne \u00d7 generation time)",
    title = "Bayesian Skyline Plot"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey88")
  )

p_skyline
```



## Clade 2


```{r}
#| label: user-settings

setwd("~/Objective1_BEAST/Objective_1_tree_annotations")
 
tree_file     <- "Data/Clade2_GTR_strict_BICEPS_comb_500_annotated.trees"    # your TreeAnnotator output file

# Your tip labels are structured as: isolate_ID|SBI|Island|Source|date
# List the field names in that exact order here:
tip_label_fields <- c("isolate_ID", "SBI", "Island", "Source", "date")
tip_label_sep    <- "|"

year_gridline_every <- 5           # gridline spacing, in years
output_plot   <- "time_tree.pdf"   # output figure filename
```

## Read the tree

Reads the tree while keeping all BEAST/TreeAnnotator annotations (posterior,
height_95%_HPD, rate, etc.) attached to each node.

```{r}
#| label: read-tree

beast_tree <- read.beast(tree_file)
beast_tree
```

Check which annotation fields are available — useful for confirming the exact
name of the HPD field before plotting (it can appear as `height_0.95_HPD` or
`height_95%_HPD` depending on TreeAnnotator settings/version):

```{r}
#| label: inspect-annotations

names(as_tibble(beast_tree))
```

## Parse metadata out of the tip labels

Since your metadata is already embedded in each tip label
(`isolate_ID|SBI|Island|Source|date`), we split on `|` to recover it as a
proper table, rather than needing a separate TSV. `Name` is kept as the
**full original tip label**, unchanged, so it still matches the tree exactly
for the join later.

```{r}
tree_tips <- beast_tree@phylo$tip.label

metadata <- tibble(Name = tree_tips) |>
  separate_wider_delim(Name, delim = tip_label_sep, names = tip_label_fields,
                        cols_remove = FALSE, too_many = "merge",
                        too_few = "align_start") |>
  mutate(date_parsed = dmy(date)) |>
  relocate(Name)

metadata
```

```{r}
#| label: check-date-parsing

# Flag any tip whose date field failed to parse -- fix these in the
# tip labels (or check for a different date format) before continuing.
metadata |> filter(is.na(date_parsed)) |> select(Name, date)
```

```{r}
#| label: most-recent-tip-date

most_recent_tip_date <- max(metadata$date_parsed, na.rm = TRUE)
most_recent_tip_date
```

This is auto-detected from the `date` field across all tips — double-check
it looks right (i.e. it's genuinely your youngest sample) before moving on.

## Build the plot

`mrsd` ("most recent sampling date") is what calibrates the x-axis to real
calendar time, using the node heights TreeAnnotator already calculated.

```{r}
#| label: build-plot
#| fig-width: 10
#| fig-height: 8

p <- ggtree(beast_tree, mrsd = as.character(most_recent_tip_date),
            ladderize = TRUE) %<+% metadata +   # joins parsed metadata by "Name"/tip label
  theme_tree2() +
  geom_range(range = "height_0.95_HPD", color = "steelblue", alpha = 0.4, size = 2) +
  geom_tippoint(aes(color = Source), size = 2) +
  geom_tiplab(aes(label = isolate_ID), size = 2.5, offset = 0.3) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
  theme(
    panel.grid.major.x = element_line(color = "grey85", linewidth = 0.3),
    panel.grid.minor.x = element_blank(),
    legend.position = "right"
  ) +
  labs(x = "Year", color = "Source")

p
```


``` {r}
#| label: build-annotated-tree
#| fig-width: 16
#| fig-height: 8

# ---- Base tree, built once ----
p <- ggtree(beast_tree, mrsd = as.character(most_recent_tip_date), ladderize = TRUE) %<+% metadata +
  theme_tree2() +
  # node age uncertainty (95% HPD bars) -- rename if your field is named differently
  geom_range(range = "height_0.95_HPD", color = "steelblue", alpha = 0.4, linewidth = 2) +
  geom_tippoint(aes(color = Source), size = 2) +
  # aligned tip labels with dotted guide lines back to the actual tip position
  geom_tiplab(aes(label = isolate_ID), size = 2, offset = 0.3, align = TRUE, linetype = "dotted") +
  theme(
    panel.grid.major.x = element_line(color = "grey85", linewidth = 0.3),
    panel.grid.minor.x = element_blank(),
    legend.position = "right"
  ) +
  labs(x = "Year", color = "Source")

# ---- Fixed tick/gridline positions, based on the tree's real range only ----
# (a live scales::pretty_breaks() function gets re-evaluated against the
# strip-widened panel later, which is why ticks kept extending outward)
tree_x_breaks <- scales::pretty_breaks(n = 15)(range(p$data$x, na.rm = TRUE))
p <- p + scale_x_continuous(breaks = tree_x_breaks)

# ---- Strip data: one column each, tip labels as row names ----
island_data <- metadata |> select(Name, Island) |> column_to_rownames("Name")
sbi_data    <- metadata |> select(Name, SBI)    |> column_to_rownames("Name")

# ---- Strip sizing, based on the tree's actual year-span ----
tree_width_years  <- diff(range(p$data$x, na.rm = TRUE))
offset1           <- 0.18 * tree_width_years    # space for tip labels before Island strip
strip_width_frac  <- 0.04                       # each strip = 4% of tree width
strip_width_years <- strip_width_frac * tree_width_years
gap_years         <- 0.5                        # gap between the two strips
offset2           <- offset1 + strip_width_years + gap_years

# ---- Custom colour palettes -- edit to match your actual categories ----
island_colors <- c(
  "NI" = "#1b9e77",
  "SI" = "#d95f02"
)

sbi_colors <- c(
  "WY12a"  = "darkred",
  "Y2c"     = "lightblue",
  "SY2c"   = "#4daf4a",
  "Y12a" = "green",
  "Y1a"  = "gold", 
  "Y1"  = "violet",
)

# ---- Strip 1: Island ----
p2 <- p + new_scale_fill()
p2 <- gheatmap(
  p2, island_data,
  offset = offset1, width = strip_width_frac,
  colnames_position = "top", colnames_angle = 90,
  colnames_offset_y = 1.5, font.size = 3
) +
  scale_fill_manual(values = island_colors, name = "Island")

# ---- Strip 2: SBI ----
p2 <- p2 + new_scale_fill()
p2 <- gheatmap(
  p2, sbi_data,
  offset = offset2, width = strip_width_frac,
  colnames_position = "top", colnames_angle = 90,
  colnames_offset_y = 1.5, font.size = 3
) +
  scale_fill_manual(values = sbi_colors, name = "SBI")

# ---- Clamp the displayed x-axis range so it doesn't blow out into the far future ----
p2 <- p2 + coord_cartesian(
  xlim = c(min(p$data$x, na.rm = TRUE) - 5, max(p$data$x, na.rm = TRUE) + 25),
  expand = FALSE
)

p2

```

## Save the figure

```{r}
#| label: save-plot

ggsave(output_plot, plot = p, width = 10, height = 8, units = "in")
message("Saved plot to: ", output_plot)
```


## User settings

```{r}
#| label: skyline-user-settings

skyline_file <- "Data/Clade2_BS_analysis_2014.712.txt"   # your exported Tracer skyline text file
```

## Read the data

The file has a title row before the real header (`time  date  datetime
milliseconds  mean  median  upper  lower`), so we skip the first line.

```{r}
#| label: read-skyline

skyline <- read_tsv(skyline_file, skip = 1, show_col_types = FALSE)
skyline
```

```{r}
#| label: inspect-skyline

glimpse(skyline)
```

## Plot

`time` is already a decimal calendar year (e.g. `2014.696`), so it can be
used directly on the x-axis without further conversion. Population size is
plotted on a log scale, which is standard for skyline plots since Ne can
span orders of magnitude.

```{r}
#| label: plot-skyline
#| fig-width: 9
#| fig-height: 6

p_skyline <- ggplot(skyline, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = median), color = "steelblue4", linewidth = 0.8) +
  scale_y_log10(labels = scales::comma) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 12)) +
  labs(
    x = "Year",
    y = "Effective population size (log scale)",
    title = "Bayesian Skyline Plot"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey88")
  )

p_skyline

p_skyline <- ggplot(skyline, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = median), color = "steelblue4", linewidth = 0.8) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 12)) +
  labs(
    x = "Year",
    y = "Effective population size (Ne \u00d7 generation time)",
    title = "Bayesian Skyline Plot"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey88")
  )

p_skyline
```





## Clade 3


```{r}
#| label: user-settings

setwd("~/Objective1_BEAST/Objective_1_tree_annotations")
 
tree_file     <- "Data/Clade3_GTR_strict_BICEPS_comb_500_annotated.trees"    # your TreeAnnotator output file

# Your tip labels are structured as: isolate_ID|SBI|Island|Source|date
# List the field names in that exact order here:
tip_label_fields <- c("isolate_ID", "SBI", "Island", "Source", "date")
tip_label_sep    <- "|"

year_gridline_every <- 5           # gridline spacing, in years
output_plot   <- "time_tree.pdf"   # output figure filename
```

## Read the tree

Reads the tree while keeping all BEAST/TreeAnnotator annotations (posterior,
height_95%_HPD, rate, etc.) attached to each node.

```{r}
#| label: read-tree

beast_tree <- read.beast(tree_file)
beast_tree
```

Check which annotation fields are available — useful for confirming the exact
name of the HPD field before plotting (it can appear as `height_0.95_HPD` or
`height_95%_HPD` depending on TreeAnnotator settings/version):

```{r}
#| label: inspect-annotations

names(as_tibble(beast_tree))
```

## Parse metadata out of the tip labels

Since your metadata is already embedded in each tip label
(`isolate_ID|SBI|Island|Source|date`), we split on `|` to recover it as a
proper table, rather than needing a separate TSV. `Name` is kept as the
**full original tip label**, unchanged, so it still matches the tree exactly
for the join later.

```{r}
tree_tips <- beast_tree@phylo$tip.label

metadata <- tibble(Name = tree_tips) |>
  separate_wider_delim(Name, delim = tip_label_sep, names = tip_label_fields,
                        cols_remove = FALSE, too_many = "merge",
                        too_few = "align_start") |>
  mutate(date_parsed = dmy(date)) |>
  relocate(Name)

metadata
```

```{r}
#| label: check-date-parsing

# Flag any tip whose date field failed to parse -- fix these in the
# tip labels (or check for a different date format) before continuing.
metadata |> filter(is.na(date_parsed)) |> select(Name, date)
```

```{r}
#| label: most-recent-tip-date

most_recent_tip_date <- max(metadata$date_parsed, na.rm = TRUE)
most_recent_tip_date
```

This is auto-detected from the `date` field across all tips — double-check
it looks right (i.e. it's genuinely your youngest sample) before moving on.

## Build the plot

`mrsd` ("most recent sampling date") is what calibrates the x-axis to real
calendar time, using the node heights TreeAnnotator already calculated.


``` {r}
#| label: build-annotated-tree
#| fig-width: 16
#| fig-height: 8

# ---- Base tree, built once ----
p <- ggtree(beast_tree, mrsd = as.character(most_recent_tip_date), ladderize = TRUE) %<+% metadata +
  theme_tree2() +
  # node age uncertainty (95% HPD bars) -- rename if your field is named differently
  geom_range(range = "height_0.95_HPD", color = "steelblue", alpha = 0.4, linewidth = 2) +
  geom_tippoint(aes(color = Source), size = 2) +
  # aligned tip labels with dotted guide lines back to the actual tip position
  geom_tiplab(aes(label = isolate_ID), size = 2, offset = 0.3, align = TRUE, linetype = "dotted") +
  theme(
    panel.grid.major.x = element_line(color = "grey85", linewidth = 0.3),
    panel.grid.minor.x = element_blank(),
    legend.position = "right"
  ) +
  labs(x = "Year", color = "Source")

# ---- Fixed tick/gridline positions, based on the tree's real range only ----
# (a live scales::pretty_breaks() function gets re-evaluated against the
# strip-widened panel later, which is why ticks kept extending outward)
tree_x_breaks <- scales::pretty_breaks(n = 15)(range(p$data$x, na.rm = TRUE))
p <- p + scale_x_continuous(breaks = tree_x_breaks)

# ---- Strip data: one column each, tip labels as row names ----
island_data <- metadata |> select(Name, Island) |> column_to_rownames("Name")
sbi_data    <- metadata |> select(Name, SBI)    |> column_to_rownames("Name")

# ---- Strip sizing, based on the tree's actual year-span ----
tree_width_years  <- diff(range(p$data$x, na.rm = TRUE))
offset1           <- 0.18 * tree_width_years    # space for tip labels before Island strip
strip_width_frac  <- 0.04                       # each strip = 4% of tree width
strip_width_years <- strip_width_frac * tree_width_years
gap_years         <- 0.5                        # gap between the two strips
offset2           <- offset1 + strip_width_years + gap_years

# ---- Custom colour palettes -- edit to match your actual categories ----
island_colors <- c(
  "NI" = "#1b9e77",
  "SI" = "#d95f02"
)

sbi_colors <- c(
  "ASY2c"  = "#e41a1c",
  "AY"     = "#377eb8",
  "SY2c"   = "#4daf4a",
  "ASWY2c" = "#984ea3",
  "AS12c"  = "yellow",
  "AS2c"   = "orange",
  "WY2c"   = "pink",
  "ASY12c" = "darkblue",
  # ---- this clade's actual values ----
  "AY2a"   = "#f781bf",
  "AWY2a"  = "#a65628",
  "A2a"    = "#999999"
)

# ---- Strip 1: Island ----
p2 <- p + new_scale_fill()
p2 <- gheatmap(
  p2, island_data,
  offset = offset1, width = strip_width_frac,
  colnames_position = "top", colnames_angle = 90,
  colnames_offset_y = 1.5, font.size = 3
) +
  scale_fill_manual(values = island_colors, name = "Island")

# ---- Strip 2: SBI ----
p2 <- p2 + new_scale_fill()
p2 <- gheatmap(
  p2, sbi_data,
  offset = offset2, width = strip_width_frac,
  colnames_position = "top", colnames_angle = 90,
  colnames_offset_y = 1.5, font.size = 3
) +
  scale_fill_manual(values = sbi_colors, name = "SBI")

# ---- Clamp the displayed x-axis range so it doesn't blow out into the far future ----
p2 <- p2 + coord_cartesian(
  xlim = c(min(p$data$x, na.rm = TRUE) - 5, max(p$data$x, na.rm = TRUE) + 25),
  expand = FALSE
)

p2

```

## Save the figure

```{r}
#| label: save-plot

ggsave(output_plot, plot = p, width = 10, height = 8, units = "in")
message("Saved plot to: ", output_plot)
```


## User settings

```{r}
#| label: skyline-user-settings

skyline_file <- "Data/Clade3_BS_analysis_2014.671.txt"   # your exported Tracer skyline text file
```

## Read the data

The file has a title row before the real header (`time  date  datetime
milliseconds  mean  median  upper  lower`), so we skip the first line.

```{r}
#| label: read-skyline

skyline <- read_tsv(skyline_file, skip = 1, show_col_types = FALSE)
skyline
```

```{r}
#| label: inspect-skyline

glimpse(skyline)
```

## Plot

`time` is already a decimal calendar year (e.g. `2014.696`), so it can be
used directly on the x-axis without further conversion. Population size is
plotted on a log scale, which is standard for skyline plots since Ne can
span orders of magnitude.

```{r}
#| label: plot-skyline
#| fig-width: 9
#| fig-height: 6

p_skyline <- ggplot(skyline, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = median), color = "steelblue4", linewidth = 0.8) +
  scale_y_log10(labels = scales::comma) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 12)) +
  labs(
    x = "Year",
    y = "Effective population size (log scale)",
    title = "Bayesian Skyline Plot"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey88")
  )

p_skyline

p_skyline <- ggplot(skyline, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = median), color = "steelblue4", linewidth = 0.8) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 12)) +
  labs(
    x = "Year",
    y = "Effective population size (Ne \u00d7 generation time)",
    title = "Bayesian Skyline Plot"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey88")
  )

p_skyline
```