---
title: "Obj_1_tree_annotations"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Overview

This notebook reads TreeAnnotator output trees (with all embedded BEAST
annotations) for three clades, calibrates each x-axis to calendar time using
the most recent tip's sampling date, parses metadata out of the tip labels,
optionally overrides that metadata with an updated external file via a
left join, and produces a time-scaled tree figure with node-age uncertainty
bars and two custom-coloured annotation strips (Island, SBI) per clade. Each
clade also gets its own Bayesian Skyline Plot section.

**Every chunk label is now unique per clade** (suffixed `-clade1` /
`-clade2` / `-clade3`) — this was required, not just tidy: Quarto errors on
duplicate chunk labels within one document, which the previous version had
throughout (three clades all reusing labels like `read-tree`,
`build-annotated-tree`, `save-plot`, etc.).

Edit each clade's **User settings** chunk, then run the whole document
(`Render` or `Ctrl/Cmd + Shift + K`).

## Setup

```{r}
#| label: load-packages

library(treeio)
library(ggtree)
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(readr)
library(ggplot2)
library(scales)
library(ggnewscale)
library(tibble)
```

## Updated metadata (optional, shared across clades)

If you have an external file with corrected/updated Island, SBI, or Source
values, load it here once. It gets left-joined onto each clade's
tip-label-derived metadata by `isolate_ID`, with **new values taking
priority and falling back to the tip-label value when the new file has no
entry for that isolate**.

**Edit the path, `read_csv`/`read_tsv` choice, and column names below to
match your actual file** — this is currently a scaffold since the exact
structure of your updated file hasn't been confirmed yet.

```{r}
#| label: load-updated-metadata

use_updated_metadata <- TRUE   # now enabled

updated_metadata <- read_csv("Data/Genome_summaries_144.csv", show_col_types = FALSE) |>
  # keep only the columns we actually want, and rename the join key to match
  # metadata's isolate_ID (note the file's own "SBI" column is used here,
  # NOT "Automated SBI" / "Automated_SBI")
  select(isolate_ID = `Isolate ID`, Island, SBI) |>
  # blanks in this file are also empty strings, not real NA -- convert so
  # coalesce() below correctly falls back to the tip-label value
  mutate(across(c(Island, SBI), ~ na_if(.x, "")))
```

A small helper so this join logic isn't repeated three times:

```{r}
#| label: apply-updated-metadata-fn

apply_updated_metadata <- function(metadata) {
  if (!use_updated_metadata) return(metadata)

  metadata |>
    left_join(updated_metadata, by = "isolate_ID", suffix = c("_old", "_new")) |>
    mutate(
      Island = coalesce(Island_new, Island_old),
      SBI    = coalesce(SBI_new, SBI_old)
      # Source isn't in updated_metadata, so it's left untouched (no _old/_new
      # suffix gets created for it by the join)
    ) |>
    select(-ends_with("_old"), -ends_with("_new"))
}
```

# Clade 1

## User settings

```{r}
#| label: user-settings-clade1

setwd("~/Objective1_BEAST/Objective_1_tree_annotations")

tree_file <- "Data/Clade1_GTR_strict_BICEPS_comb_500_annotated.trees"

tip_label_fields <- c("isolate_ID", "SBI", "Island", "Source", "date")
tip_label_sep    <- "|"

year_gridline_every <- 5
output_plot          <- "time_tree_clade1.pdf"
output_skyline_plot  <- "bayesian_skyline_plot_clade1.pdf"
skyline_file         <- "Data/Clade1_BS_analysis_2014.696.txt"
```

## Read the tree

```{r}
#| label: read-tree-clade1

beast_tree <- read.beast(tree_file)
beast_tree
```

```{r}
#| label: inspect-annotations-clade1

names(as_tibble(beast_tree))
```

## Parse metadata out of the tip labels

```{r}
#| label: parse-tip-labels-clade1

tree_tips <- beast_tree@phylo$tip.label

metadata <- tibble(Name = tree_tips) |>
  separate_wider_delim(Name, delim = tip_label_sep, names = tip_label_fields,
                        cols_remove = FALSE, too_many = "merge",
                        too_few = "align_start") |>
  mutate(
    across(all_of(tip_label_fields), ~ na_if(.x, "")),
    date_parsed = dmy(date)
  ) |>
  relocate(Name) |>
  apply_updated_metadata()

metadata

metadata |>
  summarise(across(all_of(tip_label_fields), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "field", values_to = "n_missing")
```

```{r}
#| label: check-date-parsing-clade1

metadata |> filter(is.na(date_parsed)) |> select(Name, date)
```

```{r}
#| label: most-recent-tip-date-clade1

most_recent_tip_date <- max(metadata$date_parsed, na.rm = TRUE)
most_recent_tip_date
```

## Build the annotated tree

```{r}
#| label: build-annotated-tree-clade1
#| fig-width: 16
#| fig-height: 8

p <- ggtree(beast_tree, mrsd = as.character(most_recent_tip_date), ladderize = TRUE) %<+% metadata +
  theme_tree2() +
  geom_range(range = "height_0.95_HPD", color = "steelblue", alpha = 0.4, linewidth = 2) +
  geom_tippoint(aes(color = Source), size = 2) +
  geom_tiplab(aes(label = isolate_ID), size = 2, offset = 0.3, align = TRUE, linetype = "dotted") +
  theme(
    panel.grid.major.x = element_line(color = "grey85", linewidth = 0.3),
    panel.grid.minor.x = element_blank(),
    legend.position = "right"
  ) +
  labs(x = "Year", color = "Source")

tree_x_breaks <- scales::pretty_breaks(n = 15)(range(p$data$x, na.rm = TRUE))
p <- p + scale_x_continuous(breaks = tree_x_breaks)

island_data <- metadata |>
  mutate(Island = ifelse(is.na(Island), "Unknown", Island)) |>
  select(Name, Island) |>
  column_to_rownames("Name")
sbi_data <- metadata |> select(Name, SBI) |> column_to_rownames("Name")

tree_width_years  <- diff(range(p$data$x, na.rm = TRUE))
offset1           <- 0.18 * tree_width_years
strip_width_frac  <- 0.04
strip_width_years <- strip_width_frac * tree_width_years
gap_years         <- 0.5
offset2           <- offset1 + strip_width_years + gap_years

island_colors <- c(
  "NI" = "#a6cee3",
  "SI" = "#b2df8a",
  "Unknown" = "#ffffff"
)

sbi_colors <- c(
  "AY2a" = "#1F77B4", "ASY2c" = "#AEC7E8", "WY12a" = "#FF7F0E", "Y12a" = "#FFBB78",
  "AS2c" = "#2CA02C", "SY2c" = "#98DF8A", "A2a" = "#D62728", "AWY2a" = "#FF9896",
  "Y1a" = "#9467BD", "AS12c" = "#C5B0D5", "ASWY2c" = "#8C564B", "ASY" = "#C49C94",
  "ASY12c" = "#E377C2", "AWY12a" = "#F7B6D2", "AY" = "#7F7F7F", "SWY2c" = "#C7C7C7",
  "WY1a" = "#BCBD22", "Y2c" = "#DBDB8D"
)

p2 <- p + new_scale_fill()
p2 <- gheatmap(
  p2, island_data,
  offset = offset1, width = strip_width_frac,
  colnames_position = "top", colnames_angle = 90,
  colnames_offset_y = 1.5, font.size = 3
) +
  scale_fill_manual(values = island_colors, name = "Island")

p2 <- p2 + new_scale_fill()
p2 <- gheatmap(
  p2, sbi_data,
  offset = offset2, width = strip_width_frac,
  colnames_position = "top", colnames_angle = 90,
  colnames_offset_y = 1.5, font.size = 3
) +
  scale_fill_manual(values = sbi_colors, name = "SBI")

p2 <- p2 + coord_cartesian(
  xlim = c(min(p$data$x, na.rm = TRUE) - 5, max(p$data$x, na.rm = TRUE) + 25),
  expand = FALSE
)

p2
```

## Save the figure

```{r}
#| label: save-plot-clade1

n_tips <- length(beast_tree@phylo$tip.label)
fig_height <- max(8, n_tips * 0.18)

ggsave(output_plot, plot = p2, width = 16, height = fig_height, units = "in", limitsize = FALSE)
message("Saved plot to: ", output_plot, " (height: ", round(fig_height, 1), " in)")
```

## Bayesian Skyline Plot

```{r}
#| label: read-skyline-clade1

skyline <- read_tsv(skyline_file, skip = 1, show_col_types = FALSE)
skyline
```

```{r}
#| label: inspect-skyline-clade1

glimpse(skyline)
```

```{r}
#| label: plot-skyline-clade1
#| fig-width: 9
#| fig-height: 6

p_skyline <- ggplot(skyline, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = median), color = "steelblue4", linewidth = 0.8) +
  scale_y_log10(labels = scales::comma) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 12)) +
  labs(
    x = "Year",
    y = "Effective population size (Ne \u00d7 generation time, log scale)",
    title = "Bayesian Skyline Plot — Clade 1"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey88")
  )

p_skyline
```

```{r}
#| label: save-skyline-clade1

ggsave(output_skyline_plot, plot = p_skyline, width = 9, height = 6, units = "in")
message("Saved plot to: ", output_skyline_plot)
```

# Clade 2

## User settings

```{r}
#| label: user-settings-clade2

setwd("~/Objective1_BEAST/Objective_1_tree_annotations")

tree_file <- "Data/Clade2_GTR_strict_BICEPS_comb_500_annotated.trees"

tip_label_fields <- c("isolate_ID", "SBI", "Island", "Source", "date")
tip_label_sep    <- "|"

year_gridline_every <- 5
output_plot          <- "time_tree_clade2.pdf"
output_skyline_plot  <- "bayesian_skyline_plot_clade2.pdf"
skyline_file         <- "Data/Clade2_BS_analysis_2014.712.txt"
```

## Read the tree

```{r}
#| label: read-tree-clade2

beast_tree <- read.beast(tree_file)
beast_tree
```

```{r}
#| label: inspect-annotations-clade2

names(as_tibble(beast_tree))
```

## Parse metadata out of the tip labels

```{r}
#| label: parse-tip-labels-clade2

tree_tips <- beast_tree@phylo$tip.label

metadata <- tibble(Name = tree_tips) |>
  separate_wider_delim(Name, delim = tip_label_sep, names = tip_label_fields,
                        cols_remove = FALSE, too_many = "merge",
                        too_few = "align_start") |>
  mutate(
    across(all_of(tip_label_fields), ~ na_if(.x, "")),
    date_parsed = dmy(date)
  ) |>
  relocate(Name) |>
  apply_updated_metadata()

metadata

metadata |>
  summarise(across(all_of(tip_label_fields), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "field", values_to = "n_missing")
```

```{r}
#| label: check-date-parsing-clade2

metadata |> filter(is.na(date_parsed)) |> select(Name, date)
```

```{r}
#| label: most-recent-tip-date-clade2

most_recent_tip_date <- max(metadata$date_parsed, na.rm = TRUE)
most_recent_tip_date
```

## Build the annotated tree

```{r}
#| label: build-annotated-tree-clade2
#| fig-width: 16
#| fig-height: 8

p <- ggtree(beast_tree, mrsd = as.character(most_recent_tip_date), ladderize = TRUE) %<+% metadata +
  theme_tree2() +
  geom_range(range = "height_0.95_HPD", color = "steelblue", alpha = 0.4, linewidth = 2) +
  geom_tippoint(aes(color = Source), size = 2) +
  geom_tiplab(aes(label = isolate_ID), size = 2, offset = 0.3, align = TRUE, linetype = "dotted") +
  theme(
    panel.grid.major.x = element_line(color = "grey85", linewidth = 0.3),
    panel.grid.minor.x = element_blank(),
    legend.position = "right"
  ) +
  labs(x = "Year", color = "Source")

tree_x_breaks <- scales::pretty_breaks(n = 15)(range(p$data$x, na.rm = TRUE))
p <- p + scale_x_continuous(breaks = tree_x_breaks)

island_data <- metadata |>
  mutate(Island = ifelse(is.na(Island), "Unknown", Island)) |>
  select(Name, Island) |>
  column_to_rownames("Name")
sbi_data <- metadata |> select(Name, SBI) |> column_to_rownames("Name")

tree_width_years  <- diff(range(p$data$x, na.rm = TRUE))
offset1           <- 0.18 * tree_width_years
strip_width_frac  <- 0.04
strip_width_years <- strip_width_frac * tree_width_years
gap_years         <- 0.5
offset2           <- offset1 + strip_width_years + gap_years

island_colors <- c(
  "NI" = "#a6cee3",
  "SI" = "#b2df8a",
  "Unknown" = "#ffffff"
)

sbi_colors <- c(
  "AY2a" = "#1F77B4", "ASY2c" = "#AEC7E8", "WY12a" = "#FF7F0E", "Y12a" = "#FFBB78",
  "AS2c" = "#2CA02C", "SY2c" = "#98DF8A", "A2a" = "#D62728", "AWY2a" = "#FF9896",
  "Y1a" = "#9467BD", "AS12c" = "#C5B0D5", "ASWY2c" = "#8C564B", "ASY" = "#C49C94",
  "ASY12c" = "#E377C2", "AWY12a" = "#F7B6D2", "AY" = "#7F7F7F", "SWY2c" = "#C7C7C7",
  "WY1a" = "#BCBD22", "Y2c" = "#DBDB8D"
)

p2 <- p + new_scale_fill()
p2 <- gheatmap(
  p2, island_data,
  offset = offset1, width = strip_width_frac,
  colnames_position = "top", colnames_angle = 90,
  colnames_offset_y = 1.5, font.size = 3
) +
  scale_fill_manual(values = island_colors, name = "Island")

p2 <- p2 + new_scale_fill()
p2 <- gheatmap(
  p2, sbi_data,
  offset = offset2, width = strip_width_frac,
  colnames_position = "top", colnames_angle = 90,
  colnames_offset_y = 1.5, font.size = 3
) +
  scale_fill_manual(values = sbi_colors, name = "SBI")

p2 <- p2 + coord_cartesian(
  xlim = c(min(p$data$x, na.rm = TRUE) - 5, max(p$data$x, na.rm = TRUE) + 25),
  expand = FALSE
)

p2
```

## Save the figure

```{r}
#| label: save-plot-clade2

n_tips <- length(beast_tree@phylo$tip.label)
fig_height <- max(8, n_tips * 0.18)

ggsave(output_plot, plot = p2, width = 16, height = fig_height, units = "in", limitsize = FALSE)
message("Saved plot to: ", output_plot, " (height: ", round(fig_height, 1), " in)")
```

## Bayesian Skyline Plot

```{r}
#| label: read-skyline-clade2

skyline <- read_tsv(skyline_file, skip = 1, show_col_types = FALSE)
skyline
```

```{r}
#| label: inspect-skyline-clade2

glimpse(skyline)
```

```{r}
#| label: plot-skyline-clade2
#| fig-width: 9
#| fig-height: 6

p_skyline <- ggplot(skyline, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = median), color = "steelblue4", linewidth = 0.8) +
  scale_y_log10(labels = scales::comma) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 12)) +
  labs(
    x = "Year",
    y = "Effective population size (Ne \u00d7 generation time, log scale)",
    title = "Bayesian Skyline Plot — Clade 2"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey88")
  )

p_skyline
```

```{r}
#| label: save-skyline-clade2

ggsave(output_skyline_plot, plot = p_skyline, width = 9, height = 6, units = "in")
message("Saved plot to: ", output_skyline_plot)
```

# Clade 3

## User settings

```{r}
#| label: user-settings-clade3

setwd("~/Objective1_BEAST/Objective_1_tree_annotations")

tree_file <- "Data/Clade3_GTR_strict_BICEPS_comb_500_annotated.trees"

tip_label_fields <- c("isolate_ID", "SBI", "Island", "Source", "date")
tip_label_sep    <- "|"

year_gridline_every <- 5
output_plot          <- "time_tree_clade3.pdf"
output_skyline_plot  <- "bayesian_skyline_plot_clade3.pdf"
skyline_file         <- "Data/Clade3_BS_analysis_2014.671.txt"
```

## Read the tree

```{r}
#| label: read-tree-clade3

beast_tree <- read.beast(tree_file)
beast_tree
```

```{r}
#| label: inspect-annotations-clade3

names(as_tibble(beast_tree))
```

## Parse metadata out of the tip labels

```{r}
#| label: parse-tip-labels-clade3

tree_tips <- beast_tree@phylo$tip.label

metadata <- tibble(Name = tree_tips) |>
  separate_wider_delim(Name, delim = tip_label_sep, names = tip_label_fields,
                        cols_remove = FALSE, too_many = "merge",
                        too_few = "align_start") |>
  mutate(
    across(all_of(tip_label_fields), ~ na_if(.x, "")),
    date_parsed = dmy(date)
  ) |>
  relocate(Name) |>
  apply_updated_metadata()

metadata

metadata |>
  summarise(across(all_of(tip_label_fields), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "field", values_to = "n_missing")
```

```{r}
#| label: check-date-parsing-clade3

metadata |> filter(is.na(date_parsed)) |> select(Name, date)
```

```{r}
#| label: most-recent-tip-date-clade3

most_recent_tip_date <- max(metadata$date_parsed, na.rm = TRUE)
most_recent_tip_date
```

## Build the annotated tree

```{r}
#| label: build-annotated-tree-clade3
#| fig-width: 16
#| fig-height: 8

p <- ggtree(beast_tree, mrsd = as.character(most_recent_tip_date), ladderize = TRUE) %<+% metadata +
  theme_tree2() +
  geom_range(range = "height_0.95_HPD", color = "steelblue", alpha = 0.4, linewidth = 2) +
  geom_tippoint(aes(color = Source), size = 2) +
  geom_tiplab(aes(label = isolate_ID), size = 2, offset = 0.3, align = TRUE, linetype = "dotted") +
  theme(
    panel.grid.major.x = element_line(color = "grey85", linewidth = 0.3),
    panel.grid.minor.x = element_blank(),
    legend.position = "right"
  ) +
  labs(x = "Year", color = "Source")

tree_x_breaks <- scales::pretty_breaks(n = 15)(range(p$data$x, na.rm = TRUE))
p <- p + scale_x_continuous(breaks = tree_x_breaks)

island_data <- metadata |>
  mutate(Island = ifelse(is.na(Island), "Unknown", Island)) |>
  select(Name, Island) |>
  column_to_rownames("Name")
sbi_data <- metadata |> select(Name, SBI) |> column_to_rownames("Name")

tree_width_years  <- diff(range(p$data$x, na.rm = TRUE))
offset1           <- 0.18 * tree_width_years
strip_width_frac  <- 0.04
strip_width_years <- strip_width_frac * tree_width_years
gap_years         <- 0.5
offset2           <- offset1 + strip_width_years + gap_years

island_colors <- c(
  "NI" = "#a6cee3",
  "SI" = "#b2df8a",
  "Unknown" = "#ffffff"
)

sbi_colors <- c(
  "AY2a" = "#1F77B4", "ASY2c" = "#AEC7E8", "WY12a" = "#FF7F0E", "Y12a" = "#FFBB78",
  "AS2c" = "#2CA02C", "SY2c" = "#98DF8A", "A2a" = "#D62728", "AWY2a" = "#FF9896",
  "Y1a" = "#9467BD", "AS12c" = "#C5B0D5", "ASWY2c" = "#8C564B", "ASY" = "#C49C94",
  "ASY12c" = "#E377C2", "AWY12a" = "#F7B6D2", "AY" = "#7F7F7F", "SWY2c" = "#C7C7C7",
  "WY1a" = "#BCBD22", "Y2c" = "#DBDB8D"
)

p2 <- p + new_scale_fill()
p2 <- gheatmap(
  p2, island_data,
  offset = offset1, width = strip_width_frac,
  colnames_position = "top", colnames_angle = 90,
  colnames_offset_y = 1.5, font.size = 3
) +
  scale_fill_manual(values = island_colors, name = "Island")

p2 <- p2 + new_scale_fill()
p2 <- gheatmap(
  p2, sbi_data,
  offset = offset2, width = strip_width_frac,
  colnames_position = "top", colnames_angle = 90,
  colnames_offset_y = 1.5, font.size = 3
) +
  scale_fill_manual(values = sbi_colors, name = "SBI")

p2 <- p2 + coord_cartesian(
  xlim = c(min(p$data$x, na.rm = TRUE) - 5, max(p$data$x, na.rm = TRUE) + 25),
  expand = FALSE
)

p2
```

## Save the figure

```{r}
#| label: save-plot-clade3

n_tips <- length(beast_tree@phylo$tip.label)
fig_height <- max(8, n_tips * 0.18)

ggsave(output_plot, plot = p2, width = 16, height = fig_height, units = "in", limitsize = FALSE)
message("Saved plot to: ", output_plot, " (height: ", round(fig_height, 1), " in)")
```

## Bayesian Skyline Plot

```{r}
#| label: read-skyline-clade3

skyline <- read_tsv(skyline_file, skip = 1, show_col_types = FALSE)
skyline
```

```{r}
#| label: inspect-skyline-clade3

glimpse(skyline)
```

```{r}
#| label: plot-skyline-clade3
#| fig-width: 9
#| fig-height: 6

p_skyline <- ggplot(skyline, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = median), color = "steelblue4", linewidth = 0.8) +
  scale_y_log10(labels = scales::comma) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 12)) +
  labs(
    x = "Year",
    y = "Effective population size (Ne \u00d7 generation time, log scale)",
    title = "Bayesian Skyline Plot — Clade 3"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey88")
  )

p_skyline
```

```{r}
#| label: save-skyline-clade3

ggsave(output_skyline_plot, plot = p_skyline, width = 9, height = 6, units = "in")
message("Saved plot to: ", output_skyline_plot)
```
